---
---
---

# Forest Loss and the Rise of Malaria Incidence Rate in Sub-Saharan Africa

## Data Preparation

```{r message=FALSE, warning=FALSE}
setwd("~/STAT 482")

library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(reshape2)
library(stringr)
library(openxlsx)
library(patchwork)
library(purrr)
library(GGally)
library(plotly)
library(scales)
library(ggcorrplot)
library(nlme)
library(readxl)
library(ggrepel)


# turn off scientific notation format output
options(scipen = 999)

# The clean_char function takes in a character and returns a number without the range in brackets and without spaces in the mean. 
# For example, "193 232 [149 000-245 000]" becomes 193232
clean_char = function(text){
  return(as.numeric(gsub(" ", "", sub("\\[.*", "", text))))
}
```

#### Population by Country

```{r}
# Import Subsaharan African countries list
countries = read_excel("~/STAT 482/subsaharan_countries.xlsx", sheet = "Countries", col_names =  TRUE)

population = read_excel("~/STAT 482/Population.xls", sheet = "Data", col_names =  TRUE)

# filter to get only sub-Saharan African countries
population = population[population$`Country Name` %in% countries$country_name,]

# keep only years 2000-2020, remove Indicator Name and Indicator Code cols.
to_remove = as.character(c(1960:1999,2021:2023, "Indicator Code", "Indicator Name"))
population = population[,!(colnames(population) %in% to_remove)]
```

#### Forest Cover Data

```{r message=FALSE, warning=FALSE}
# import forest cover data by country and the sub-Saharan countries list
forest = read.csv("forest_coverage_data_by_country.xls - Data.csv")

# take out the first two rows of forest, which contain data we don't need
forest = forest[-c(1,2),]

# make the first row of the data the headers
colnames(forest) = as.character(unlist(forest[1,]))
forest = forest[-1,]

# filter to get only sub-Saharan African countries
forest = forest[forest$`Country Name` %in% countries$country_name,]

# keep only years 2000-2020, remove Indicator Name and Indicator Code cols.
forest = forest[,!(colnames(forest) %in% to_remove)]
```

#### Malaria Case Data

```{r message=FALSE}
# import malaria case number data by country
malaria = read_xlsx("estimated_malaria_cases.xlsx")

# filter to get only Sub-saharan African countries
malaria = malaria[malaria$`Country Name` %in% countries$country_name,]


# loop through all numeric columns in malaria, applying clean_char function
col_mal = colnames(malaria) # list of column names of malaria data
for (i in 2:length(col_mal)){
  malaria[col_mal[i]] = sapply(malaria[,i], clean_char)}


# remove column for year 2021
malaria = malaria[,!(colnames(malaria) %in% to_remove)]

# create table with country name and country code for joining
country_name_code = forest[,c(1:2)] 

# join the country code to the malaria dataset by the country name
malaria = left_join(malaria, country_name_code, by = "Country Name")
```

##### Normalize Malaria Case Data by Population per Country

```{r}
# Reshape to long format
malaria_long = malaria %>%
  pivot_longer(cols = starts_with("20"), names_to = "year", values_to = "cases")

population_long = population %>%
  pivot_longer(cols = starts_with("20"), names_to = "year", values_to = "population")

# Join the data
combined_data = malaria_long %>%
  left_join(population_long, by = c("Country Name", "Country Code", "year"))

# Normalize the data
normalized_data = combined_data %>%
  mutate(normalized_cases = cases / population)

# Reshape back to wide format
malaria = normalized_data %>%
  select("Country Name", "Country Code", year, normalized_cases) %>%
  pivot_wider(names_from = year, values_from = normalized_cases)
```

#### Rainfall Data

```{r}
rainfall = read_excel("~/STAT 482/rainfall_data.xls", sheet = "Data", col_names =  TRUE) 

rainfall = rainfall[rainfall$`Country Name` %in% countries$country_name,]
rainfall = rainfall[,!(colnames(rainfall) %in% to_remove)]
```

#### GDP per Capita Data

```{r}
gdp = read_excel("~/STAT 482/GDP_per_capita_total.xls", sheet = "Data", col_names = TRUE)

gdp = gdp[gdp$`Country Name` %in% countries$country_name,]
gdp = gdp[,!(colnames(gdp) %in% to_remove)]
```

#### Urbanization Rate Data

```{r}
urbanization = read_excel("~/STAT 482/urbanization_rate_data.xls", sheet = "Data", col_names = TRUE)

urbanization = urbanization[urbanization$`Country Name` %in% countries$country_name,]
urbanization = urbanization[,!(colnames(urbanization) %in% to_remove)]
```

### Combine Data into one Dataframe

```{r}
# reorder the 'country' variable in order of most north to most south in latitude
countries_ordered = c("Eritrea", "Djibouti", "Chad", "Niger", "Mali", "Mauritania", "Sudan", "Ethiopia", "Somalia", "Central African Republic", "Cameroon", "Nigeria", "Benin", "Burkina Faso", "Ghana", "Togo", "Equatorial Guinea", "Gabon", "Congo, Rep.", "Congo, Dem. Rep.", "Uganda", "Kenya", "Rwanda", "Burundi", "Tanzania", "Angola", "Zambia", "Malawi", "Mozambique", "Zimbabwe", "Botswana", "Namibia", "South Africa", "Eswatini", "Madagascar", "Senegal", "Gambia, The", "Guinea-Bissau", "Guinea", "Sierra Leone", "Liberia", "Sao Tome and Principe")

# assign the ordered country names to malaria and forest dataset
malaria$`Country Name` = factor(malaria$`Country Name`, levels = countries_ordered)
forest$`Country Name` = factor(forest$`Country Name`, levels = countries_ordered)


# reshape datasets to long format
malaria_long = malaria %>%
  pivot_longer(cols = starts_with("2"), names_to = "Year", values_to = "Malaria_Incidence")

forest_long = forest %>%
  pivot_longer(cols = starts_with("2"), names_to = "Year", values_to = "Forest_Cover")

rainfall_long = rainfall %>%
  pivot_longer(cols = starts_with("2"), names_to = "Year", values_to = "Rainfall_Depth")

gdp_long = gdp %>%
  pivot_longer(cols = starts_with("2"), names_to = "Year", values_to = "GDP_per_Capita")

urban_long = urbanization %>%
  pivot_longer(cols = starts_with("2"), names_to = "Year", values_to = "Urbanization_Perc")

# Merge datasets
combined_data = malaria_long %>%
  left_join(forest_long, by = c("Country Name", "Year")) %>%
  left_join(rainfall_long, by = c("Country Name", "Year")) %>%
  left_join(gdp_long, by = c("Country Name", "Year")) %>%
  left_join(urban_long, by = c("Country Name", "Year"))

combine_remove = as.character(c("Country Code.x", "Country Code.x.x", "Country Code.y.y", "Country Code.y"))
combined_data = combined_data[,!(colnames(combined_data) %in% combine_remove)]
```

#### Handle NAs

```{r}

temp = combined_data

# Filter the dataset for non-missing GDP per capita values and for Eritrea
eritrea_data = temp[!is.na(temp$GDP_per_Capita) & temp$`Country Name` == "Eritrea", "GDP_per_Capita"]

# Calculate the average GDP per capita for Eritrea
eritreaGDPavg = sum(eritrea_data) / nrow(eritrea_data)

# Replace NA values with the calculated average
temp$GDP_per_Capita[is.na(temp$GDP_per_Capita) & temp$`Country Name` == "Eritrea"] = eritreaGDPavg


# Filter the dataset for non-missing Rainfall_Depth values and for Sudan
sudan_data = temp[!is.na(temp$Rainfall_Depth) & temp$`Country Name` == "Sudan", "Rainfall_Depth"]

# Calculate the average Rainfall_Depth for Sudan
sudanRainfallAvg = sum(sudan_data) / nrow(sudan_data)

# Replace NA values with the calculated average
temp$Rainfall_Depth[is.na(temp$Rainfall_Depth) & temp$`Country Name` == "Sudan"] = sudanRainfallAvg

combined_data = temp
```

------------------------------------------------------------------------

## Report 1: Exploratory Data Analysis

#### Scatter plots of Malaria Incidence and Forest Cover Percentage by Country

```{r fig.height=11, fig.width=10, message=FALSE, warning=FALSE}

# make the year numeric for plotting
combined_data$Year = as.numeric(combined_data$Year)


# forest scatterplot of Year vs Forest Cover Percentage by Country
ggplot(data = combined_data, aes(x = Year, 
                                 y = Forest_Cover, 
                                 color = `Country Name`)) +
  geom_point(size = 3) +
  labs(title = "Year vs Forest Cover Percentage", 
       x = "Year", 
       y = "Forest Cover Percentage") +
  theme(
    legend.text = element_text(size = 10),
    axis.title.x = element_text(size = 15),
    axis.title.y = element_text(size = 15), 
    plot.title = element_text(size = 20), 
    axis.text = element_text(size = 9)
  ) +
  guides(color = guide_legend(ncol = 1, title = NULL))


# malaria scatterplot of Year vs Malaria Incidence by Country
ggplot(data = combined_data, aes(x = Year, 
                                 y = Malaria_Incidence, 
                                 color = `Country Name`)) +
  geom_point(size = 3) +
  labs(title = "Year vs Malaria Incidence", 
       x = "Year", 
       y = "Malaria Incidence") +
  theme(
    legend.text = element_text(size = 10),
    axis.title.x = element_text(size = 15),
    axis.title.y = element_text(size = 15), 
    plot.title = element_text(size = 20), 
    axis.text = element_text(size = 9)
  ) +
  guides(color = guide_legend(ncol = 1, title = NULL))
```

#### Box plots of Malaria Incidence and Forest Cover Percentage by Country (one plot)

```{r fig.height=10, fig.width=11, message=FALSE, warning=FALSE}

# boxplot of malaria incidence by country - all in one plot
ggplot(data = combined_data, aes(x = `Country Name`, y = Malaria_Incidence)) +
  geom_boxplot(fill = "lightblue", color = "black", outlier.colour = "red", outlier.shape = 16) +
  labs(title = "Boxplot of Malaria Incidence by Country", x = "Country", y = "Malaria Incidence") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 20),
    axis.title.x = element_text(size = 15),
    axis.title.y = element_text(size = 15),
    axis.text = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1))

# boxplot of forest cover percentage by country - all in one plot
ggplot(data = combined_data, aes(x = `Country Name`, y = Forest_Cover)) +
  geom_boxplot(fill = "lightblue", color = "black", outlier.colour = "red", outlier.shape = 16) +
  labs(title = "Boxplot of Forest Cover Percentage by Country", x = "Country", y = "Forest Cover Percentage") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 20),
    axis.title.x = element_text(size = 15),
    axis.title.y = element_text(size = 15),
    axis.text = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1))
```

#### Heat map of Malaria Incidence and Forest Cover Percentage by Country

```{r fig.height=10, fig.width=10}

# format forest cover data for heatmap compatability
forest_long = melt(combined_data, id.vars = c("Year", "Country Name"), 
                           measure.vars = "Forest_Cover")

ggplot(forest_long, aes(x = Year, y = `Country Name`, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "forest green") +
  labs(title = "Heatmap of Forest Cover Percentage by Year and Country", 
       x = "Year", 
       y = "Country", 
       fill = "Forest Cover Percentage") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 20),
    axis.title.x = element_text(size = 15),
    axis.title.y = element_text(size = 15),
    axis.text = element_text(size = 12))


# format malaria data for heatmap compatability
malaria_long = melt(combined_data, id.vars = c("Year", "Country Name"), 
                           measure.vars = "Malaria_Incidence")

ggplot(malaria_long, aes(x = Year, y = `Country Name`, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "red") +
  labs(title = "Heatmap of Malaria Incidence by Year and Country", 
       x = "Year", 
       y = "Country", 
       fill = "Malaria Incidence") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 20),
    axis.title.x = element_text(size = 15),
    axis.title.y = element_text(size = 15),
    axis.text = element_text(size = 12))
```

#### Scatterplots by Country - Malaria and Forest Cover per Year

```{r fig.height=70, fig.width=14, message=FALSE, warning=FALSE}
# Scatter plot for all countries and years

# Split the data by country
split_data = combined_data %>% split(.$`Country Name`)

# Create a list of plots for each country
plots_scatter = map2(split_data, names(split_data), ~ {
  ggplot(.x, aes(x = Year)) +
    geom_line(aes(y = `Forest_Cover`, color = "Forest Cover %"), linewidth = 1) +
    geom_line(aes(y = `Malaria_Incidence` * 100, color = "Malaria Incidence"), linewidth = 1) +  # Scale Malaria Incidence
    scale_y_continuous(
      name = "Forest Cover %",
      sec.axis = sec_axis(~ . / 100, name = "% of Population with Malaria")  # Adjust the secondary axis
    ) +
    labs(title = paste("Forest Cover % and % of Population with Malaria in", .y)) +
    theme(legend.position = "bottom") +
    scale_color_manual(
      name = "Legend",
      values = c("Forest Cover %" = "Forest Green", "Malaria Incidence" = "red")
    )
})

# Combine all plots into one display with free scales
combined_plot_scatter = wrap_plots(plots_scatter, ncol = 2)

# Display the combined plot
print(combined_plot_scatter)
```

#### Scatterplots by Country - All Covariates and Response

```{r fig.height=200, fig.width=15, message=FALSE, warning=FALSE}
# Function to create side-by-side histograms for a single country
create_histograms = function(country_data, country_name) {
  malaria_plot = ggplot(country_data, aes(x = Year, y = `Malaria_Incidence`)) +
    geom_col(fill = "red", alpha = 0.7) +
    labs(title = paste("% of Population with Malaria in", country_name), x = "Year", y = "% of Population with Malaria") +
    theme_minimal()
  
  forest_plot = ggplot(country_data, aes(x = Year, y = `Forest_Cover`)) +
    geom_col(fill = "Forest Green", alpha = 0.7) +
    labs(title = paste("Forest Cover % in", country_name), x = "Year", y = "Forest Cover %") +
    theme_minimal()
  
  rainfall_plot = ggplot(country_data, aes(x = Year, y = `Rainfall_Depth`)) +
    geom_col(fill = "blue", alpha = 0.7) +
    labs(title = paste("Rainfall Depth in", country_name), x = "Year", y = "Rainfall Depth mm^3") +
    theme_minimal()
  
  gdp_plot = ggplot(country_data, aes(x = Year, y = `GDP_per_Capita`)) +
    geom_col(fill = "yellow", alpha = 0.7) +
    labs(title = paste("GDP per Capita in", country_name), x = "Year", y = "GDP per Capita, $US") +
    theme_minimal()
  
  urbanization_plot = ggplot(country_data, aes(x = Year, y = `Urbanization_Perc`)) +
    geom_col(fill = "orange", alpha = 0.7) +
    labs(title = paste("Urbanization Percentage in", country_name), x = "Year", y = "Urbanization %") +
    theme_minimal()
  
  malaria_plot + forest_plot + rainfall_plot + gdp_plot + urbanization_plot
}

# Create a list of plots for each country
plots = combined_data %>%
  split(.$`Country Name`) %>%
  map2(names(.), ~ create_histograms(.x, .y))

# Combine all plots into one display
combined_plot = wrap_plots(plots, ncol = 1)

# Display the combined plot
print(combined_plot)
```

#### Correlation Matrix

```{r fig.height=5, fig.width=5}

# Get only data columns and rename them for accurate data labels
corr_data = combined_data[,c("Malaria_Incidence", "Forest_Cover", "Rainfall_Depth", "Urbanization_Perc", "GDP_per_Capita")]
colnames(corr_data) = c("% Population with Malaria", "Forest Cover %", "Rainfall Depth mm^3", "Urbanization %", "GDP per Capita $US")

# Compute correlation matrix
cor_matrix = cor(corr_data)

# Create the correlation matrix plot
ggcorrplot(cor_matrix, 
           method = "circle", 
           type = "full", 
           lab = TRUE, 
           tl.cex = 8,
           colors = c("red", "white", "forest green"),
           title = "Correlation Matrix of Various Factors", ggtheme = theme_gray())
```

------------------------------------------------------------------------

## Report 2 - Linear Regression

#### Centralize Covariates

```{r}
# Centralize the Covariate to have beta0 be the average malaria rate for an average country
covariates = c("Forest_Cover", "Urbanization_Perc", "Rainfall_Depth", "GDP_per_Capita")

# Centralize the covariates
for (covariate in covariates) {
  combined_data[[paste0(covariate)]] = combined_data[[covariate]] - mean(combined_data[[covariate]], na.rm = TRUE)
}
```

#### Fit Linear Regression Models for All Years

Includes model summaries and Normal QQ-Plots

```{r}
# Fit a linear regression model including all year and country data
fit_overall = lm(Malaria_Incidence ~ Forest_Cover + Urbanization_Perc + Rainfall_Depth + GDP_per_Capita, data = combined_data)
qq_norm_res=qqnorm(residuals(fit_overall),main = "Normal Q-Q Plot - Overall")
qqnorm_model=lm(qq_norm_res$y~qq_norm_res$x)
abline(a=qqnorm_model$coef[1], b=qqnorm_model$coef[2])
summary(fit_overall)

# Add Year (numerical) as a covariate in the overall model
fit_overall_w_year = lm(Malaria_Incidence ~ Year + Forest_Cover + Urbanization_Perc + Rainfall_Depth + GDP_per_Capita, data = combined_data)
qq_norm_res=qqnorm(residuals(fit_overall_w_year),main = "Normal Q-Q Plot - Overall with Year Covariate")
qqnorm_model=lm(qq_norm_res$y~qq_norm_res$x)
abline(a=qqnorm_model$coef[1], b=qqnorm_model$coef[2])
summary(fit_overall_w_year)


# Print summary of coefficients and R^2
cat("\n", "Overall Model", "\n")
print(fit_overall$coefficients)
cat("\n", "R^2:", summary(fit_overall)$r.squared, "\n", "R^2 adjusted:", summary(fit_overall)$adj.r.squared, "\n")
cat("\n", "Overall Model with Year Covariate", "\n")
print(fit_overall_w_year$coefficients)
cat("\n", "R^2:", summary(fit_overall_w_year)$r.squared, "\n", "R^2 adjusted:", summary(fit_overall_w_year)$adj.r.squared, "\n")
```

#### Fit Linear Regression Model per Year

```{r}
# Fit the model per year

# List to store models
models_list = list()
years = unique(combined_data$Year)

# Fit the model for each year and store in the list
for (year in years) {
  data_year = subset(combined_data, Year == year)
  model = lm(Malaria_Incidence ~ Forest_Cover + Urbanization_Perc + Rainfall_Depth + GDP_per_Capita, data = data_year)
  models_list[[as.character(year)]] = summary(model)
}

# Extract p-values, standard errors, estimates, and t-statistics
p_values = data.frame()
standard_errors = data.frame()
estimates = data.frame()
t_statistics = data.frame()

for (year in names(models_list)) {
  model_summary = models_list[[year]]
  coefs = coef(model_summary)
  
  p_values[year, names(coefs[, "Pr(>|t|)"])] = coefs[, "Pr(>|t|)"]
  standard_errors[year, names(coefs[, "Std. Error"])] = coefs[, "Std. Error"]
  estimates[year, names(coefs[, "Estimate"])] = coefs[, "Estimate"]
  t_statistics[year, names(coefs[, "t value"])] = coefs[, "t value"]
}

p_values$Year = c(2000:2020)
standard_errors$Year = c(2000:2020)
estimates$Year = c(2000:2020)
t_statistics$Year = c(2000:2020)

# Convert to long format for plotting
p_values_long = melt(p_values, id.vars = "Year", variable.name = "Variable", value.name = "P_Value")
standard_errors_long = melt(standard_errors, id.vars = "Year", variable.name = "Variable", value.name = "Standard_Error")
estimates_long = melt(estimates, id.vars = "Year", variable.name = "Variable", value.name = "Estimate")
t_statistics_long = melt(t_statistics, id.vars = "Year", variable.name = "Variable", value.name = "T_Value")

```

##### Plot Statistics from Models per Year

```{r}
# Plot the statistics from models per year

# Define a color palette for the variables including the intercept
variable_colors = c(
  "(Intercept)" = "#9467bd",
  "Forest_Cover" = "#1f77b4",
  "Urbanization_Perc" = "#ff7f0e",
  "Rainfall_Depth" = "#2ca02c",
  "GDP_per_Capita" = "#d62728"
)

# Plotting p-values
ggplot(p_values_long, aes(x = Year, y = P_Value, color = Variable)) +
  geom_line() +
  geom_point() +
  scale_color_manual(values = variable_colors) +
  labs(title = "P-Values Over Years", y = "P-Value")


# Plotting estimates for each variable
for (variable in unique(estimates_long$Variable)) {
  plot = ggplot(subset(estimates_long, Variable == variable), aes(x = Year, y = Estimate, color = Variable)) +
    geom_line() +
    geom_point() +
    scale_color_manual(values = variable_colors) +
    labs(title = paste("Estimates Over Years for", variable), y = "Estimate") +
    theme_minimal()
  
  print(plot)  # Display the plot
}
```

#### Goodness-of-Fit for Models per Year

```{r}
# Perform Goodness-of-Fit and plot results

# List to store models
models_list = list()

# Unique years in the data
years = unique(combined_data$Year)

# Data frame to store goodness-of-fit metrics
goodness_of_fit = data.frame(Year = integer(), p_value = numeric(), Adjusted_R_squared = numeric())

# Fit the model for each year and store the goodness-of-fit metrics
for (year in years) {
  data_year = subset(combined_data, Year == year)
  
  # Fit the model
  model = lm(Malaria_Incidence ~ Forest_Cover + Urbanization_Perc + Rainfall_Depth + GDP_per_Capita, data = data_year)
  
  # Get the summary of the model
  model_summary = summary(model)
  
  # Perform the F-test
  f_statistic = model_summary$fstatistic["value"]
  numdf = model_summary$fstatistic["numdf"]
  dendf = model_summary$fstatistic["dendf"]
  p_value = pf(f_statistic, numdf, dendf, lower.tail = FALSE)
  
  # Extract Adjusted R-squared
  adjusted_r_squared = model_summary$adj.r.squared
  
  # Store the metrics in the data frame
  goodness_of_fit = rbind(goodness_of_fit, data.frame(Year = year, p_value = p_value, Adjusted_R_squared = adjusted_r_squared))
}

# Print the goodness-of-fit metrics
print(goodness_of_fit)

# Plot p-value over the years
ggplot(goodness_of_fit, aes(x = Year, y = p_value)) +
  geom_line() +
  geom_point() +
  labs(title = "P-Value Over Years", x = "Year", y = "P-Value") +
  theme_minimal()

# Plot Adjusted R-squared over the years
ggplot(goodness_of_fit, aes(x = Year, y = Adjusted_R_squared)) +
  geom_line() +
  geom_point() +
  labs(title = "Adjusted R-Squared Over Years", x = "Year", y = "Adjusted R-Squared") +
  theme_minimal()
```

## Report 3: Mixed Effects Model

#### Model Selection

```{r warning=FALSE}
# DATA PREP
# Rename 'Country Name' to 'country_name'
mixed_data = combined_data %>%
  dplyr::rename(country_name = `Country Name`)

# Convert 'country_name' to a factor
mixed_data$country_name = as.factor(mixed_data$country_name)

# FIT MODELS
# only random forest cover
fit.nlme1 = lme(Malaria_Incidence ~ Rainfall_Depth + Urbanization_Perc + Forest_Cover, 
                 random = ~ Forest_Cover | country_name, 
                 data = mixed_data, 
                 control = lmeControl(opt = "optim", maxIter = 1000, msMaxIter = 1000))


# random forest cover and rainfall depth
fit.nlme2 = lme(Malaria_Incidence ~ Rainfall_Depth + Urbanization_Perc + Forest_Cover, 
                 random = ~ Rainfall_Depth + Forest_Cover | country_name, 
                 data = mixed_data, 
                 control = lmeControl(opt = "optim", maxIter = 1000, msMaxIter = 1000))


# remove rainfall depth from model, only forest cover is random 
fit.nlme3 = lme(Malaria_Incidence ~ Urbanization_Perc + Forest_Cover + Year, 
                 random = ~ Forest_Cover | country_name, 
                 data = mixed_data, 
                 control = lmeControl(opt = "optim", maxIter = 1000, msMaxIter = 1000))


# Compare models
anova(fit.nlme1, fit.nlme2, fit.nlme3)


summary(fit.nlme1) # only forest cover is random
summary(fit.nlme2) # forest cover and rainfall depth are random
summary(fit.nlme3) # remove rainfall depth from model, only forest cover random
```

#### Model Visualizations

```{r warning=FALSE}
# Create visuals for the model with only Forest_Cover as the random effect

# Extract residuals from the model
residuals = residuals(fit.nlme3)

# Create QQ plot
qqnorm(residuals)
qqline(residuals, col = "red")



# Extract fixed effects coefficients and their confidence intervals
fixed_effects <- summary(fit.nlme3)$tTable
fixed_effects_df <- data.frame(
  Term = rownames(fixed_effects),
  Estimate = fixed_effects[, "Value"],
  Std_Error = fixed_effects[, "Std.Error"],
  t_value = fixed_effects[, "t-value"],
  P_Value = fixed_effects[, "p-value"]
)
fixed_effects_df$CI_Lower <- fixed_effects_df$Estimate - 1.96 * fixed_effects_df$Std_Error
fixed_effects_df$CI_Upper <- fixed_effects_df$Estimate + 1.96 * fixed_effects_df$Std_Error

# Remove the intercept from the data frame
fixed_effects_df <- fixed_effects_df[fixed_effects_df$Term != "(Intercept)", ]

# Plot fixed effects coefficients with their confidence intervals
ggplot(fixed_effects_df, aes(x = Term, y = Estimate)) +
  geom_point() +
  geom_errorbar(aes(ymin = CI_Lower, ymax = CI_Upper), width = 0.2) +
  labs(title = "Fixed Effects Coefficients with Confidence Intervals", x = "Estimate", y = "Value") +
  theme_minimal()





# Extract random effects for each country
random_effects <- ranef(fit.nlme3)
random_effects_df <- as.data.frame(random_effects)

# Rename columns to avoid issues with special characters
colnames(random_effects_df) <- c("Random_Intercept", "Random_Slope_Forest_Cover")

# Plot random effects for each country with non-overlapping text labels
ggplot(random_effects_df, aes(x = Random_Intercept, y = Random_Slope_Forest_Cover, label = rownames(random_effects_df))) +
  geom_point() +
  geom_text_repel(size = 2.5) +
  labs(title = "Random Effects for Each Country", x = "Random Intercept", y = "Random Slope for Forest Cover") +
  theme_minimal()




# Plot residuals vs fitted values
residuals <- resid(fit.nlme3)
fitted_values <- fitted(fit.nlme3)
residuals_df <- data.frame(Fitted_Values = fitted_values, Residuals = residuals)

ggplot(residuals_df, aes(x = Fitted_Values, y = Residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted Values", x = "Fitted Values", y = "Residuals") +
  theme_minimal()
```
